Refinement for open automata

### 2021-03-22 2021-Refinement/rapport.tex
Refinement v0b
	modified:   rapport.pdf
	modified:   rapport.tex

\usepackage{xunicode}
\usepackage{fontspec}
\usepackage{geometry}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{unicode-math}
\AtBeginDocument{\renewcommand\setminus{\smallsetminus}}
\usepackage{polyglossia}
\setmainlanguage{english}

@@

\newcommand\nmm[1]{\(\displaystyle #1\)}
\newcommand\setR{\mathbb{R}}
\newcommand\setZ{\mathbb{Z}}
\newcommand\setN{\mathbb{N}}
\newcommand\mpar[1]{{\left(#1\right)}}
\newcommand\mbrk[1]{{\left[#1\right]}}
\newcommand\mbrc[1]{{\left\{#1\right\}}}
\newcommand\mdbrk[1]{\left⟦#1\right⟧}
\newcommand\card[1]{{\left|#1\right|}}
\newcommand\psubst[1]{\mbrc{\!\!\mbrc{#1}\!\!}}
\newcommand\subst[2]{\mbrk{#1\middle/#2}}
\newcommand\choice[1]{\left\{\everymath{\displaystyle}%
	\begin{array}{lr}#1\end{array}\right.}
\newcommand\midbar{\,\middle|\,}
\newcommand\mset[2]{\mbrc{#1\midbar #2}}
\newcommand\prel[4]{\mbrk{#2\mathrel{#1}#3\midbar #4}}
\newcommand\subbox[1]{{\makebox[.5\width]{\(\scriptstyle #1\)}}}
\newcommand\bigsymb[2][\Large]{\text{#1\nmm{#2}}}
\newcommand\defnotation{\DOTSB\;{\Colon=}\;}
\newcommand\defobject{\DOTSB\;{\coloneq}\;}
\newcommand\nwedge{\DOTSB\;{\wedge}\;}
\newcommand\qwedge{\DOTSB\quad{\wedge}\quad}
\newcommand\sat{{\vdash}}
\newcommand\fvars[1]{{\mathit{vars}\mpar{#1}}}
\newcommand\fterms[1]{{\mathcal{T}\mpar{#1}}}
\newcommand\fcterms[1]{{\mathcal{T}_{\makebox[1pt][r]{\(\scriptstyle 0\)}}\mpar{#1}}}
\newcommand\fformulas[1]{{\mathcal{F}\mpar{#1}}}
\newcommand\fcformulas[1]{{\mathcal{F}_{\makebox[1pt][r]{\(\scriptstyle 0\)}}\mpar{#1}}}
\newcommand\fvalues[1]{{\mathcal{P}\mpar{#1}}}
\newcommand\OA[7]{\left<#1,#2,#3,#4, #5,#6,#7\right>}
\newcommand\OT[6]{\text{\small\(%
	\setlength\arraycolsep{2pt}\everymath{\displaystyle}%
	\begin{array}{c}%
	#4,#5,#6 \\\hline
	#1\xrightarrow{\raisebox{-1.5pt}[.8\height][0pt]{\makebox[1.4\width]{\(\scriptstyle #3\)}}}#2
	\end{array}\)}}
\newcommand\OTx[4]{\OT{s_{#1}}{s'_{#1 #2}}{\alpha_{#1 #2}}{\beta_{#3 j}^{j \in J'_{#4}}}{g_{#1 #2}}{\psi_{#1 #2}}}
\newcommand\OTg{\OTx{}{}{}{}}

@@

\section{Open Automaton}
To define an open automaton we need some preliminary definitions.
\begin{defi}[Expression algebra]
An expression algebra \(E\) is a disjoint union \(E \defobject \mathcal{T} \uplus \mathcal{F}\) where \(\mathcal{T}\) is a term algebra and \(\mathcal{F}\) the formulas over \(\mathcal{T}\).

%TODO: there might be binders in expressions, vars = free veriables, quantifiers are also allowed
We also define convenience notations \(\fterms{E} \defnotation \mathcal{T}\) and \(\fformulas{E} \defnotation \mathcal{F}\).

@@

The formulas form a propositional logic over terms in \(\mathcal{T}\) and some relations, or equivalently they are first order logic formulas without quantifiers but free variables are allowed.
An example of term algebra can be Peano integers (zero, variable and successor function), the formulas associated can use the relations equality, \(\mathit{sum}\mpar{a, b, c} \defobject a = b + c\) and \(\mathit{prod}\mpar{a, b, c} \defobject a = b \times c\).
\begin{defi}[Variables, Closed terms/formulas]
\(\fvars{e}\) is the set of variables in \(e \in E\).

@@

The closed formulas are \(\fcformulas{E} \defnotation \mset{f \in \fformulas{E}}{\fvars{f} = \emptyset}\); \(\fcformulas{E} \subset \fformulas{E}\).

@@

More generally a closed expression is an expression without variable.
In a standard context it would be without free variables but here the ``no quantifier" constraint means that there is no such thing as ``bound variables".

@@

\item The satisfaction of closed formulas, \(\sat \in \fcformulas{E} \to \mbrc{0, 1}\);

@@

% \item The formula interpretation of any variable assignement \(\psi \in \fterms{E}^V\), \nmm{\mdbrk{\psi} \defnotation \bigwedge_{v \in V} v = \psi\mpar{v}}; \(\mdbrk{\psi} \in \fformulas{E}\).

@@

The interpretation of values and satisfaction are supposed to be decidable.

@@

\item Let \nmm{f \in \fcformulas{E}}, \nmm{\vdash f_0 \defnotation \sat\mpar{f_0} = 1} and \nmm{\nvdash f_0 \defnotation \sat\mpar{f_0} = 0};
\item Let \nmm{f \in \fformulas{E}, \sigma \in \fvalues{E}^V}, \\
	\nmm{\sigma \vdash f \defnotation \exists \varphi \in \fvalues{E}^{\fvars{f} \setminus V}, f\psubst{\sigma \uplus \varphi} \in \fcformulas{E} \nwedge \vdash f\psubst{\sigma \uplus \varphi}}.

@@

% TODO: redo that explaination

@@

The second notation means that the disjunction of the right formulas is satisfiable, not necessarily with the same value for a variable name between formulas, for all valid valuation of the left formula.
\begin{defi}[Formulas combination, Expression algebra combination]
The propositional logic built without existential quantifiers on top of two propositional logics \(\fformulas{E_1}\) and \(\fformulas{E_2}\) is \(\fformulas{E_1} + \fformulas{E_2}\).
By extension \(E_1 + E_2 \defnotation \mpar{\fformulas{E_1} + \fformulas{E_2}} \uplus \mpar{\fterms{E_1} \cup \fterms{E_2}}\).

With these common definitions and notations settled, the objects of interest can now be defined.
\begin{defi}[Open automaton]
An open automaton is a tuple \(\OA{S}{i}{E}{V}{\varphi}{J}{T}\) with \(S\) the set of states, \(i \in S\) the initial state, \(E\) an expression algebra, \(V\) the set of variable names unique to this automaton, \(\varphi \in \fvalues{E}^V\) the initial valuation of variables, \(J\) the set of hole names and \(T\) a set of open-transitions.

\(S, V, J\) are arbitrary sets, only \(J\) is required to be finite.
\end{defi}
The variable names may clash when considering two automaton, in that case we suppose that we can still distinguish the variables.
All the possible open transitions, from which transitions in \(T\) are selected, are defined below.
\begin{defi}[Open transition]
An open transition is a tuple \nmm{\OT{s}{s'}{\alpha\mpar{v_\alpha}}{\mpar{\beta_j\mpar{v_{\beta j}}}^{j \in J'}}{g}{\psi}} with \(s, s' \in S\) the source and target states, \(\alpha \in A\) an action label, \(J' \subseteq J\) the holes involved, \(\beta_j \in A\) the action labels of the holes, \(v_\alpha, v_{\beta j}^{j \in J'}\) fresh distinct variables, \(g \in \fformulas{E}\) the guard and \(\psi \in \fterms{E}^V\) the new value of the variables.

The set of action labels \(A\) is arbitrary and implicit.
To simplify formulas, the notations \(\alpha \defnotation \alpha\mpar{v_\alpha}\) and \(\alpha = \beta \defnotation \alpha = \beta \wedge v_\alpha = v_\beta\) will be used.
The guard and new values of the variables can use any defined variable: \nmm{\bigcup_{v \in V} \fvars{\psi\mpar{v}} \cup \fvars{g} \subseteq V \cup \mbrc{v_\alpha} \cup \mset{v_{\beta j}}{j \in J'}}.
\end{defi}
Now we can define some utilitary functions:
\begin{defi}[Guard, Variables, Out-transition, In-transition]
\(\mathrm{OT}\mpar{s}\) are called the out-transitions of \(s\) and \(\mathrm{IT}\mpar{s}\) and its in-transitions.
\begin{align*}
	\mathit{guard}\mpar{\OTg} & \defnotation g & \mathit{}
	\fvars{\OT{s}{s'}{\alpha\mpar{v_\alpha}}{\mpar{\beta_j\mpar{v_{\beta j}}}^{j \in J'}}{g}{\psi}} & \defnotation \mbrc{v_\alpha} \cup \mset{v_{\beta j}}{j \in J'} \\
	\mathrm{OT}\mpar{s} & \defnotation \mbrc{\OTg \in T} &
	\mathrm{IT}\mpar{s'} & \defnotation \mbrc{\OTg \in T}
\end{align*}
\end{defi}
The intuition of an open automaton is a partially defined LTS with variables, guards on transitions and parametrised actions.

\section{Semantic and composition of Open Automata} % This is taken from 2007.10770.pdf
% Intro section
% \begin{defi}[Run of an open autmaton]
% A run of an open automaton \(\mathrm{OA}_c \defobject \OA{S}{i}{E}{V}{\varphi}{J}{T}\) is a list \(\rho\) of pairs \(\forall k, \rho\mbrk{k} \defobject \mpar{t_k, \nu_k}\) with \(t_k \in T, \nu_k \in \fcterms{E}^\fvars{t_k}\) such that:

% Let \nmm{t_k \defnotation \OTx{k}{}{k}{k}},
% \begin{itemize}
% \item \(\forall k, s_k = \choice{i & k = 0 \\ s'_{k-1} & k > 0}\);
% \item \(\forall k, \sigma\mpar{\rho}\mbrk{k} \defnotation \choice{\varphi & k = 0 \\ \sigma\mpar{\rho}\mbrk{k-1}\psubst{\psi_{k-1}}\psubst{\nu_{k-1}} & k > 0}\), \(\sigma\mpar{\rho}\mbrk{k} \in \fvalues{E}^V\);
% \item \(\vdash g_k\psubst{\sigma\mpar{\rho}\mbrk{k} \uplus \nu_k}\).
% \end{itemize}
% \end{defi}
% Explain what a run is and what purpose each element serves.
When the auomaton is in a state \(s \in S\) with a valuation of its variables \(\sigma \in \fvalues{E}^V\), it cannot perform transitions \nmm{t \defobject \OTg \in T} such that \(\nvdash g\psubst{\sigma}\).
After performing the transition and emitting action \(\alpha\), the automaton is in the state \(s'\) with valuation \(\psi\psubst{\sigma \uplus \varphi}\) for some \(\varphi \in \fvalues{E}^\fvars{t}\).
The holes are synchronised using a handshake like mecanism when the transition happens.
\begin{exi} % Examples
\end{exi}

An open automaton can be partially specified, the partial specification comes from the holes.
A hole can be filled with an open automaton, this operation is called composition.
\begin{defi}[Composition of open automata]
The composition of \(\mathrm{OA}_c \defobject \OA{S_c}{i_c}{E_c}{V_c}{\varphi_c}{J_c}{T_c}\) in the hole \(k \in J_p\) of \(\mathrm{OA}_p \defobject \OA{S_p}{i_p}{E_p}{V_p}{\varphi_p}{J_p}{T_p}\) is:
\begin{align*}
	\mathrm{OA}_p\subst{\mathrm{OA}_c}{k} \defnotation & \OA{S_p \times S_c}{\mpar{i_p, i_c}}{E_p + E_c}{V_p \uplus V_c}{\varphi_p \uplus \varphi_c}{J_c \uplus J_p \setminus \mbrc{k}}{T} \\
	\text{With } T \defobject & \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s'_c}}{\alpha_p}{\beta_j^{j \in J'_c \uplus J'_p \setminus \mbrc{k}}}{g_p \wedge g_c \wedge \alpha_c = \beta_k}{\psi_p \uplus \psi_c}}{\OTx{p}{}{}{p} \in T_p, \OTx{c}{}{}{c} \in T_c} \\
	& \cup \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s_c}}{\alpha_p}{\beta_j^{j \in J'_p}}{g_p}{\psi_p}}{\OTx{p}{}{}{p}, k \notin J', s_c \in S_c}
\end{align*}
Similarly to expression substitution, the parallel composition is noted \(\mathrm{OA}\psubst{\mathrm{OA}_j^{j \in J}}\).
\end{defi}
The actions emitted when \(\mathrm{OA}_c\) makes a transition is sychronised with the action of the hole \(k\) in transitions of \(\mathrm{OA}_p\) which have it as a hole actions (first transition set, \(\alpha_c = \beta_k\)).
No transition of \(\mathrm{OA}_c\) is performed when a transition that do not refer to the hole \(k\) is performed in \(\mathrm{OA}_p\) (second transition set, \(k \notin J'\)).
The composition may look like a handshake, with both automata running in parallel (product of states and joint variables) but it is actually not symmetric.
\begin{exi} % TODO: Examples
\end{exi}

@@

The important properties we will consider here are:
\begin{description}
\item[Reflexivity:] \(\forall a, a \leq a\);
\item[Transitivity:] \(\forall a\, b\, c, a \leq b \wedge b \leq c \implies a \leq c\);
\item[Preorder:] Reflexivity and transitivity;
\item[Composition is a refinement:] \(\forall a\, b, a\mbrk{b} \leq a\);
\item[Composition context equivalence:] \(\forall a\, b\, c, a \leq b \wedge a\mbrk{c} \leq a \implies a\mbrk{c} \leq b\mbrk{c}\);
\item[Composition congruence:] \(\forall a\, b\, c, a \leq b \wedge c\mbrk{a} \leq c \implies c\mbrk{a} \leq c\mbrk{b}\);
\item[Composition compatibility:] \(\forall a\, b\, c\, d, a \leq b \wedge c \leq d \wedge c\mbrk{a} \leq c \implies c\mbrk{a} \leq d\mbrk{b}\);
\item[FH-bisimulation compatibility:] \(\forall a\, b\, c, d, a = b \wedge c = d \wedge a \leq c \implies b \leq d\).
\end{description}
The interesting but not necessary properties are:
\begin{description}
\item[Deadlock preservation:] If \(a \leq b\) then \(a\) does not introduces new deadlocks.
\end{description}
Deadlock preservation conflicts with ``composition is a refinement" by disallowing some unwanted cases where an automaton in a hole cannot produce any action that out-transitions expect.
In particular filling a hole with the deadlock automaton (one state without out-transitions) is not considered a refinement for a deadlock preserving refinement.
This property has another impact which lead to write prerequisites like \(a \leq b \wedge a\mbrk{c} \leq a \implies a\mbrk{c} \leq b\mbrk{c}\) for context equivalence because \(a\mbrk{c} \leq a\) might not be true.

On the other hand the interesting but actually unwanted properties are:
\begin{description}
\item[Daisy equivalence:] Filling a hole of \(a\) with a daisy gives \(b\) such that \(a \leq b \wedge b \leq a\), i.e. it is not a strict refinement.
\end{description}
While daisy equivalence might seem to be a natural property because the daisy being the meaning of a hole, there is actually a difference between the two.
Daisy equivalence allows to refine by filling a daisy but also by going the other way, that is putting new action restrictions to any transition (and duplicating transitions).
The issue especially arise when two independant holes are filled with a daisy, then the other way is simulated by putting back the actions but as a unique hole, effectively merging independant holes.
% TODO: Expand on why not to do that

% refinement on Open Automata
% Caveat (is blablabla a strict refinement) and what we may want (trace/run safety)
% Working definition next section

\section{Refinement relation}
The main objective of the refinement relation in this section is to be able to say that a composition is a refinement of the base automaton (\(a\mbrk{b} \leq a\)).
However composing an open automaton can give an automaton where not much can be said in terms of hole indicies.
So looking at restricted cases where holes are related in a specific manner can help understanding the general case.
For two open automata \(\mathrm{OA}_1 \defobject \OA{S_1}{i_1}{E_1}{J_1}{V_1}{\varphi_1}{T_1}\) and \(\mathrm{OA}_2 \defobject \OA{S_2}{i_2}{E_2}{J_2}{V_2}{\varphi_2}{T_2}\), the following refinement relations are defined.

This relation is the basis for all the following ones.
It applies on automaton with same holes.
The goal is to caracterise automata that have a more determined behaviour (more deterministic) without adding deadlocks.
\begin{defi}[Hole-identical refinement]
If \(J_1 = J_2\) then ``\(\mathrm{OA}_1\) is a hole-identical refinement of \(\mathrm{OA}_2\)" is defined, and its definition is as follows:
\begin{multline*}
	\mathrm{OA}_1 \leq_\equiv \mathrm{OA}_2 \defnotation \exists R \in \mpar{S_1 \times S_2} \to \fformulas{E}_1 + \fformulas{E}_2, \\
	\mpar{\exists p \in \fformulas{E}_1 + \fformulas{E}_2, \quad \prel{R}{i_1}{i_2}{p} \qwedge \vdash p\psubst{\varphi_1 \uplus \varphi_2}} \\
	\wedge R \text{ hole-identical simulation of } \mathrm{OA}_1 \text{ in } \mathrm{OA}_2
\end{multline*}
\end{defi}
As in other simulation-like relations, \(R\) can be seen as a witness of \(\mathrm{OA}_1 \leq_\equiv \mathrm{OA}_2\).
Hence when two open automata are in relation, such a simulation will be called a witness.
To be a witness the standard requirement is that \(R\) relates initial states, \(R\) is closed under progress and related states are simultaneously final.

>>>>

\usepackage{xunicode}
\usepackage{fontspec}
\usepackage{geometry}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{unicode-math}
\usepackage{xcolor}
\AtBeginDocument{\renewcommand\setminus{\smallsetminus}}
\usepackage{polyglossia}
\setmainlanguage{english}

@@

\newcommand\comment[3]{\colorbox{#1}{#2}\marginpar{#3}}
\newcommand\Rabea{\comment{yellow}}
\newcommand\Ludo{\comment{green}}
\newcommand\Eric{\comment{cyan}}
\newcommand\Quentin{\comment{pink}}

\newcommand\nmm[1]{\(\displaystyle #1\)} % nmm for Nice Math Mode
\newcommand\setR{\mathbb{R}}
\newcommand\setZ{\mathbb{Z}}
\newcommand\setN{\mathbb{N}}
\newcommand\mpar[1]{{\left(#1\right)}}
\newcommand\mbrk[1]{{\left[#1\right]}}
\newcommand\mbrc[1]{{\left\{#1\right\}}}
\newcommand\mdbrk[1]{\left⟦#1\right⟧}
\newcommand\card[1]{{\left|#1\right|}}
\newcommand\psubst[1]{\mbrc{\!\!\mbrc{#1}\!\!}}
\newcommand\subst[2]{\mbrk{#1\middle/#2}}
\newcommand\choice[1]{\left\{\everymath{\displaystyle}%
	\begin{array}{lr}#1\end{array}\right.}
\newcommand\midbar{\,\middle|\,}
\newcommand\mset[2]{\mbrc{#1\midbar #2}}
\newcommand\prel[4]{\mbrk{#2\mathrel{#1}#3\midbar #4}}
\newcommand\subbox[1]{{\makebox[.5\width]{\(\scriptstyle #1\)}}}
\newcommand\bigsymb[2][\Large]{\text{#1\nmm{#2}}}
\newcommand\defnotation{\DOTSB\;{\Colon=}\;}
\newcommand\defobject{\DOTSB\;{\coloneq}\;}
\newcommand\nwedge{\DOTSB\;{\wedge}\;} % when you want to force space
\newcommand\qwedge{\DOTSB\quad{\wedge}\quad}
\newcommand\sat{{\vdash}} % suppress spaces
\newcommand\fvars[1]{{\mathit{vars}\mpar{#1}}}
\newcommand\fterms[1]{{\mathcal{T}}}
\newcommand\fcterms[1]{{\mathcal{T}_{\makebox[1pt][r]{\(\scriptstyle 0\)}}}}
\newcommand\fformulae[1]{{\mathcal{F}}}
\newcommand\fcformulae[1]{{\mathcal{F}_{\makebox[1pt][r]{\(\scriptstyle 0\)}}}}
% \newcommand\fterms[1]{{\mathcal{T}\mpar{#1}}}
% \newcommand\fcterms[1]{{\mathcal{T}_{\makebox[1pt][r]{\(\scriptstyle 0\)}}\mpar{#1}}}
% \newcommand\fformulae[1]{{\mathcal{F}\mpar{#1}}}
% \newcommand\fcformulae[1]{{\mathcal{F}_{\makebox[1pt][r]{\(\scriptstyle 0\)}}\mpar{#1}}}
% \newcommand\fvalues[1]{{\mathcal{P}\mpar{#1}}}
\newcommand\fvalues[1]{{\mathcal{P}}}
\newcommand\OA[6]{\left<#1,#2,#3,#4,#5,#6\right>}
\newcommand\OT[6]{\text{\small\(%
	\setlength\arraycolsep{2pt}\everymath{\displaystyle}%
	\begin{array}{c}%
	#4,#5,#6 \\\hline
	#1\xrightarrow{\raisebox{-1.5pt}[.8\height][0pt]{\makebox[1.4\width]{\(\scriptstyle #3\)}}}#2
	\end{array}\)}}
\newcommand\OTx[4]{\OT{s_{#1}}{s'_{#1 #2}}{\alpha_{#1 #2}}{\beta_{#3 j}^{j \in J'_{#4}}}{g_{#1 #2}}{\psi_{#1 #2}}}
\newcommand\OTg{\OTx{}{}{}{}}

@@

\section{Open Automaton}
To define an open automaton we need some preliminary definitions.
\begin{defi}[Expression algebra]
An expression algebra \(E\) is a disjoint union \(E \defobject \mathcal{T} \uplus \mathcal{F}\) where \(\mathcal{T}\) is a term algebra and \(\mathcal{F}\) the formulae over \(\mathcal{T}\).
% We also define convenience notations \(\fterms{E} \defnotation \mathcal{T}\) and \(\fformulae{E} \defnotation \mathcal{F}\).

@@

The formulae form at least a propositional logic over terms in \(\mathcal{T}\) and some relations, or equivalently they are at least first order logic formulae without quantifiers but with free variables allowed.
An example of term algebra can be Peano integers (zero, variable and successor function), the formulae associated can use the relations equality, \(\mathit{sum}\mpar{a, b, c} \defobject a = b + c\) and \(\mathit{prod}\mpar{a, b, c} \defobject a = b \times c\).
\begin{defi}[Free variables, Closed terms/formulae]
\(\fvars{e}\) is the set of unbound variables in \(e \in E\).

@@

The closed formulae are \(\fcformulae{E} \defnotation \mset{f \in \fformulae{E}}{\fvars{f} = \emptyset}\); \(\fcformulae{E} \subset \fformulae{E}\).

@@

More generally a closed expression is an expression without unbound variable.
If the expression algebra has no binding construction then closed terms are terms without variables.

@@

\item The satisfaction of closed formulae, \(\sat \subseteq \fcformulae{E} \to \mbrc{0, 1, ?}\);

@@

% \item The formula interpretation of any variable assignement \(\psi \in \fterms{E}^V\), \nmm{\mdbrk{\psi} \defnotation \bigwedge_{v \in V} v = \psi\mpar{v}}; \(\mdbrk{\psi} \in \fformulae{E}\).

@@

The interpretation of values is supposed to be decidable.
The satisfaction of formulae is not supposed to be decidable, however it is better if it is ;).

@@

\item Let \nmm{f \in \fcformulae{E}}, \nmm{\vdash f_0 \defnotation \sat\mpar{f_0} = 1} and \nmm{\nvdash f_0 \defnotation \sat\mpar{f_0} = 0};
\item Let \nmm{f \in \fformulae{E}, \sigma \in \fvalues{E}^V}, \\
	\nmm{\sigma \vdash f \defnotation \exists \varphi \in \fvalues{E}^{\fvars{f} \setminus V}, f\psubst{\sigma \uplus \varphi} \in \fcformulae{E} \nwedge \vdash f\psubst{\sigma \uplus \varphi}}.

@@

@@

The second notation is the satisfiability of formulae under a (partial) valuation.
% \begin{defi}[Formulas combination, Expression algebra combination]
% The propositional logic built without existential quantifiers on top of two propositional logics \(\fformulae{E_1}\) and \(\fformulae{E_2}\) is \(\fformulae{E_1} + \fformulae{E_2}\).
% By extension \(E_1 + E_2 \defnotation \mpar{\fformulae{E_1} + \fformulae{E_2}} \uplus \mpar{\fterms{E_1} \cup \fterms{E_2}}\).
% \end{defi}

From there the expression algebra is fixed.
With these common definitions and notations settled, the objects of interest can now be defined.
\begin{defi}[Open automaton]
An open automaton is a tuple \(\OA{S}{i}{V}{\varphi}{J}{T}\) with \(S\) the set of states, \(i \in S\) the initial state, \(V\) the set of variable names unique to this automaton, \(\varphi \in \fvalues{E}^V\) the initial valuation of variables, \(J\) the set of hole names and \(T\) a set of open-transitions.

\(S, V, J\) are arbitrary sets, \Quentin{only \(J\) is required to be finite.}{Is \(S\) also required to be finite?}
\end{defi}
The variable names may clash when considering two automata, in that case we suppose that we can still distinguish the variables.
All the possible open transitions, from which transitions in \(T\) are selected, are defined below.
\begin{defi}[Open transition]
An open transition is a tuple \nmm{\OT{s}{s'}{\alpha\mpar{v_\alpha}}{\mpar{\beta_j\mpar{v_{\beta j}}}^{j \in J'}}{g}{\psi}} with \(s, s' \in S\) the source and target states, \(\alpha \in A\) an action label, \(J' \subseteq J\) the holes involved, \(\beta_j \in A\) the action labels of the holes, \(v_\alpha, v_{\beta j}^{j \in J'}\) fresh distinct variables, \(g \in \fformulae{E}\) the guard and \(\psi \in \fterms{E}^V\) the new value of the variables.

\Quentin{The set of action labels \(A\) is arbitrary and implicit.}{For the moment}
To simplify formulae, the notations \(\alpha \defnotation \alpha\mpar{v_\alpha}\) and \(\alpha = \beta \defnotation \alpha = \beta \wedge v_\alpha = v_\beta\) will be used.
The guard and new values of the variables can use any defined variable: \nmm{\bigcup_{v \in V} \fvars{\psi\mpar{v}} \cup \fvars{g} \subseteq V \cup \mbrc{v_\alpha} \cup \mset{v_{\beta j}}{j \in J'}}.
\end{defi}
Now we can define some utilitary functions:
\begin{defi}[Guard, Variables, Out-transition, In-transition]
\(\mathrm{OT}\mpar{s}\) are called the out-transitions of \(s\) and \(\mathrm{IT}\mpar{s}\) and its in-transitions.
\begin{align*}
	\mathit{guard}\mpar{\OTg} & \defnotation g & \mathit{}
	\fvars{\OT{s}{s'}{\alpha\mpar{v_\alpha}}{\mpar{\beta_j\mpar{v_{\beta j}}}^{j \in J'}}{g}{\psi}} & \defnotation \mbrc{v_\alpha} \cup \mset{v_{\beta j}}{j \in J'} \\
	\mathrm{OT}\mpar{s} & \defnotation \mbrc{\OTg \in T} &
	\mathrm{IT}\mpar{s'} & \defnotation \mbrc{\OTg \in T}
\end{align*}
\end{defi}
The intuition of an open automaton is a partially defined LTS with variables, guards on transitions and parametrised actions.

\section{Semantic and composition of Open Automata} % This is taken from 2007.10770.pdf
% TODO: Intro
When the auomaton is in a state \(s \in S\) with a valuation of its variables \(\sigma \in \fvalues{E}^V\), it cannot perform transitions \nmm{t \defobject \OTg \in T} such that \(\nvdash g\psubst{\sigma}\).
After performing the transition and emitting action \(\alpha\), the automaton is in the state \(s'\) with valuation \(\psi\psubst{\sigma \uplus \varphi}\) for some \(\varphi \in \fvalues{E}^\fvars{t}\).
The holes are synchronised using a handshake like mecanism when the transition happens.
\begin{exi} % Examples
\end{exi}

An open automaton can be partially specified, the partial specification comes from the holes.
A hole can be filled with an open automaton, this operation is called composition.
\begin{defi}[Composition of open automata]
The composition of \(\mathrm{OA}_c \defobject \OA{S_c}{i_c}{V_c}{\varphi_c}{J_c}{T_c}\) in the hole \(k \in J_p\) of \(\mathrm{OA}_p \defobject \OA{S_p}{i_p}{V_p}{\varphi_p}{J_p}{T_p}\) is:
\begin{align*}
	\mathrm{OA}_p\subst{\mathrm{OA}_c}{k} \defnotation & \OA{S_p \times S_c}{\mpar{i_p, i_c}}{V_p \uplus V_c}{\varphi_p \uplus \varphi_c}{J_c \uplus J_p \setminus \mbrc{k}}{T} \\
	\text{With } T \defobject & \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s'_c}}{\alpha_p}{\beta_j^{j \in J'_c \uplus J'_p \setminus \mbrc{k}}}{g_p \wedge g_c \wedge \alpha_c = \beta_k}{\psi_p \uplus \psi_c}}{\OTx{p}{}{}{p} \in T_p, \OTx{c}{}{}{c} \in T_c} \\
	& \cup \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s_c}}{\alpha_p}{\beta_j^{j \in J'_p}}{g_p}{\psi_p}}{\OTx{p}{}{}{p}, k \notin J', s_c \in S_c}
\end{align*}
Similarly to expression substitution, the parallel composition is noted \(\mathrm{OA}\psubst{\mathrm{OA}_j^{j \in J}}\).
\end{defi}
The actions emitted when \(\mathrm{OA}_c\) makes a transition is sychronised with the action of the hole \(k\) in transitions of \(\mathrm{OA}_p\) which have it as a hole actions (first transition set, \(\alpha_c = \beta_k\)).
No transition of \(\mathrm{OA}_c\) is performed when a transition that do not refer to the hole \(k\) is performed in \(\mathrm{OA}_p\) (second transition set, \(k \notin J'\)).
The composition may look like a handshake, with both automata running in parallel (product of states and joint variables) but it is actually not symmetric.
\begin{exi} % TODO: Examples
\end{exi}

@@

The important properties we will consider here are:
\begin{description}
\item[Reflexivity:] \(\forall a, a \leq a\);
\item[Transitivity:] \(\forall a\, b\, c, a \leq b \wedge b \leq c \implies a \leq c\);
\item[Preorder:] Reflexivity and transitivity;
\item[Composition is a refinement:] \(\forall a\, b, a\mbrk{b} \leq a\);
\item[Composition context equivalence:] \(\forall a\, b\, c, a \leq b \wedge a\mbrk{c} \leq a \implies a\mbrk{c} \leq b\mbrk{c}\);
\item[Composition congruence:] \(\forall a\, b\, c, a \leq b \wedge c\mbrk{a} \leq c \implies c\mbrk{a} \leq c\mbrk{b}\);
\item[Composition compatibility:] \(\forall a\, b\, c\, d, a \leq b \wedge c \leq d \wedge c\mbrk{a} \leq c \implies c\mbrk{a} \leq d\mbrk{b}\);
\item[FH-bisimulation compatibility:] \(\forall a\, b\, c, d, a = b \wedge c = d \wedge a \leq c \implies b \leq d\).
\end{description}
The interesting but not necessary properties are:
\begin{description}
\item[No deadlock introduction:] If \(a \leq b\) then \(a\) does not introduces new deadlocks.
\end{description}
``No deadlock introduction" conflicts with ``composition is a refinement" by disallowing some unwanted cases where an automaton in a hole cannot produce any action that out-transitions expect.
In particular filling a hole with the deadlock automaton (one state without out-transitions) is not considered a refinement for a no deadlock introductive refinement.
This property has another impact which lead to write prerequisites like \(a \leq b \wedge a\mbrk{c} \leq a \implies a\mbrk{c} \leq b\mbrk{c}\) for context equivalence because \(a\mbrk{c} \leq a\) might not be true.

On the other hand the interesting but actually unwanted properties are:
\begin{description}
\item[Daisy equivalence:] Filling a hole of \(a\) with a \Quentin{daisy}{How is it called in english?} gives \(b\) such that \(a \leq b\) and \(b \leq a\), i.e. it is not a strict refinement.
\end{description}
While daisy equivalence might seem to be a natural property because the daisy being the meaning of a hole, there is actually a difference between the two.
Daisy equivalence allows to refine by filling a daisy but also by going the other way, that is putting new action restrictions to any transition (and duplicating transitions).
The issue especially arise when two independant holes are filled with a daisy, then the other way is simulated by putting back the actions but as a unique hole, effectively merging independant holes.
% TODO: Expand on why not to do that, using an example for instance
\begin{exi}
\end{exi}

% refinement on Open Automata
% Caveat (is blablabla a strict refinement) and what we may want (trace/run safety)
\begin{defi}[Refinement simulation correction] % Correction definition
\end{defi}
% Working definition next section

\section{Refinement relation}
The main objective of the refinement relation in this section is to be able to say that a composition is a refinement of the base automaton (\(a\mbrk{b} \leq a\)).
However composing an open automaton can give an automaton where not much can be said in terms of hole indicies.
So looking at restricted cases where holes are related in a specific manner can help understanding the general case.
For two open automata \(\mathrm{OA}_1 \defobject \OA{S_1}{i_1}{J_1}{V_1}{\varphi_1}{T_1}\) and \(\mathrm{OA}_2 \defobject \OA{S_2}{i_2}{J_2}{V_2}{\varphi_2}{T_2}\), the following refinement relations are defined.
